Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R

### 2017-04-24 meehan_et_al_inla_nmix_jss/article.tex
tweaks
incorporated haavard's comments

%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Timothy D. Meehan\\National\\Audubon Society \And
        Nicole L. Michel\\National\\Audubon Society \And
        H\"{a}vard Rue\\King Abdulla University\\of Science and Technology}
\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}

%% comma-separated
\Plainauthor{Timothy D. Meehan, Nicole L. Michel, H\"{a}vard Rue}
%% without formatting
\Plaintitle{Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R}
%% a short title
\Shorttitle{\pkg{R-INLA} for N-Mixture Models}

@@

%% The address of (at least) one author should be given
\Address{
  Timothy D. Meehan\\
  National Audubon Society\\
  Boulder, Colorado USA\\
  E-mail: \email{tmeehan@audubon.org}\\
\\
  Nicole L. Michel\\
  National Audubon Society\\
  San Francisco, California USA\\
  E-mail: \email{nmichel@audubon.org}\\
  \\
  H\"{a}vard Rue\\
  King Abdulla University
  of Science and Technology\\
  Thuwal, SAU\\
  E-mail: \email{haavard.rue@kaust.edu.sa }
}
%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}. One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area. If, during repeat visits, the population is assumed to be closed (no immigration , emigration, or reproduction; static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability. This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.

Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture model \citep{Royle_Nichols_2003, Dodd_Dorazio_2004, Royle_2004, Kery_Royle_Schmid_2005}. A simple form of an N-mixture model, a binomial mixture model, describes observed counts $Y$ at site $i$ during survey $j$ as coming from a binomial distribution with parameters for abundance $N$ and detection probability $p$, where $N$ per site is drawn from a Poisson distribution with an expected value, lambda $(\lambda)$. Specifically,

$$N_i \sim \text{Poisson}(\lambda) \qquad \text{and} \qquad  Y_{i,j} | N_i \sim \text{binomial}(N_i, p).$$

$\lambda$ is commonly modeled as a log-linear function of site covariates, as $\text{log}(\lambda_i) = \beta_0 + \beta_1 * x_i$. Similarly, $p$ is commonly modeled as $\text{logit}(p_{i,j}) = \alpha_0 + \alpha_1 * x_{i,j}$, a logit-linear function of site-survey covariates.

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for wildlife species that breed annually \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{binomial}(N_{i,k}, p_{i,k})$. Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates. There are other variations of N-mixture models that accommodate overdispersed counts through use of a negative binomial distribution \citep{Kery_Royle_2010}, a zero-inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}. Yet other variations account for non-independent detection probabilities through use of a beta-binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}. We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for an overview.

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology. However, there are practical issues that sometimes act as barriers to adoption. Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}. These are extremely powerful and flexible platforms for analyzing hierarchical models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the \proglang{BUGS} modeling syntax. While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment. Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, and may not occur at all.

@@

Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} for \proglang{R}. The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009}. INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC. Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in \proglang{R}. Also, like \pkg{unmarked}, the computational cost of analyzing models with \pkg{R-INLA} is relatively low. The \pkg{R-INLA} approach is different from \pkg{unmarked} in that inference about model parameters falls within a Bayesian framework.

@@

Much of the code used to conduct analyses is shown in the body of this manuscript. However, some repeated code, and code related to generating tables and figures, is not shown for brevity. All code can be accessed via \url{https://github.com/tmeeha/inlaNMix}.

@@

As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion was generated and modeled using a negative binomial distribution for the count component of the model. The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.b4 = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

@@

\subsection[Analysis with JAGS]{Analysis with JAGS}
We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package. In defining the model, we specified a negative binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.

@@

The first argument in the call to the pcount() function was the model formula, which specified the covariates for detection and then the covariates for abundance. This was followed by an argument identifying the appropriate unmarked frame, and the form of the mixture model, negative binomial-binomial.

@@

The call to the inla() function includes several components. First is the model statement. On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.  On the right side of the formula is a 1 to specify a global intercept for $p$ and the two covariates for $p$.  Note that a wide range of random effects for $p$ could be added to the right side of the formula using the f() syntax \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013}. The second argument describes the data, provided here as a list that corresponds with the model formula. Third is the likelihood family, which can take values of "nmix" for a Poisson-binomial mixture and "nmixnb" for a negative binomial-binomial mixture. The fourth (control.fixed, for detection parameters) and fifth (control.family, for abundance and overdispersion parameters) arguments specify the priors for the two model components. In this case, the priors are specified similarly to those in the \proglang{JAGS} model, but a variety of other options are available. Several other characteristics of the analysis can be modified in the call to inla(). See \cite{Rue_Martino_Lindgren_Simpson_Riebler_2013} for details.

@@

%% discussion
\section[Discussion]{Discussion}
The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows Frequentist inference.

@@

Limitations of \pkg{R-INLA} are mainly related to the more restricted set of N-mixture models that can be specified. Of the approaches explored here, \proglang{BUGS}-based approaches allow users ultimate flexibility in specifying models. For example, with \proglang{JAGS}, site-survey-year covariates for detection are possible, multiple types of mixed distributions are available \citep{Joseph_Elkin_Martin_Possingham_2009,Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, and a variety of random effects can be specified for both $\lambda$ and $p$ \citep{Kery_Schaub_2011}. In comparison, the current version of \pkg{R-INLA} does not handle site-survey covariates, employs only Poisson-binomial and negative binomial-binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for $p$ only. In cases where site-survey covariates are particularly important and not otherwise controlled in the sampling design, \pkg{R-INLA} will not be the appropriate tool (Fig. \ref{fig:fig2}).

When compared to \pkg{unmarked}, the \pkg{R-INLA} approach is similar in regards to model syntax and data format. The approaches are also similar in that both yield results much faster than MCMC. The two approaches differ in that \pkg{R-INLA} is roughly 10 times faster than \pkg{unmarked}, possibly due to the different approach used to compute model likelihoods (see Appendix A). They also differ in that \pkg{unmarked} can accommodate site-survey covariates and can analyze dynamic N-mixture models \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.

@@

%% appendix
\appendix
\section{Recursive computations of the "nmix" likelihood}
The likelihood for the simplest case is
\begin{displaymath}
    \text{Probability}(y) = \sum_{n = y}^{\infty}
    \text{Poisson}(n ; \lambda) \;\times\; \text{binomial}(y;  n, p)
\end{displaymath}
where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{binomial}(y; n, p)$ is the density for the binomial distribution with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$. Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we can easily evaluate it using a recursive algorithm that is both faster and more numerical stable. The same idea is also applicable to the negative binomial case, and the case where we have replicated observations of the same $n$. We leave it to the reader to derive these straight forward extensions.

The key observation is that both the Poisson and the binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{binomial}(y; n, p) = \text{binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \texttt{fac} starting with the smallest contributions, which are more numerically stable.

>>>>

%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Timothy D. Meehan\\National\\Audubon Society \And
        Nicole L. Michel\\National\\Audubon Society \And
        Haavard Rue\\King Abdulla University\\of Science and Technology}
\title{Estimating Animal Abundance with N-Mixture Models Using the \pkg{R-INLA} Package for R}

%% comma-separated
\Plainauthor{Timothy D. Meehan, Nicole L. Michel, Haavard Rue}
%% without formatting
\Plaintitle{Estimating Animal Abundance with N-Mixture Models Using the R-INLA Package for R}
%% a short title
\Shorttitle{\pkg{R-INLA} for N-Mixture Models}

@@

%% The address of (at least) one author should be given
\Address{
  Timothy D. Meehan\\
  National Audubon Society\\
  Boulder, Colorado USA\\
  E-mail: \email{tmeehan@audubon.org}\\
\\
  Nicole L. Michel\\
  National Audubon Society\\
  San Francisco, California USA\\
  E-mail: \email{nmichel@audubon.org}\\
  \\
  Haavard Rue\\
  King Abdulla University
  of Science and Technology\\
  Thuwal, Saudi Arabia\\
  E-mail: \email{haavard.rue@kaust.edu.sa }
}
%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In recent years, new sampling schemes and modeling approaches have enabled improved estimates of animal abundance that are less biased by non-detection \citep{Denes_Silveira_Beissinger_2015}. One such sampling scheme, termed a metapopulation design \citep{Kery_Royle_2010}, involves repeat visits in rapid succession to each of multiple study sites in a study area. If, during repeat visits, the population is assumed to be closed (no immigration, emigration, or reproduction; static abundance), then the ratio of detections to non-detections during repeated counts can inform an estimate of detection probability. This detection probability can, in effect, be used to correct abundance estimates for imperfect detection \citep{Royle_2004}.

Data resulting from this sampling scheme are often modeled using an explicitly hierarchical statistical model referred to as an N-mixture model \citep{Royle_Nichols_2003, Dodd_Dorazio_2004, Royle_2004, Kery_Royle_Schmid_2005}. A simple form of an N-mixture model, a Binomial mixture model, describes observed counts $Y$ at site $i$ during survey $j$ as coming from a Binomial distribution with parameters for abundance $N$ and detection probability $p$, where $N$ per site is drawn from a Poisson distribution with an expected value $\lambda$. Specifically,

$$N_i \sim \text{Poisson}(\lambda) \qquad \text{and} \qquad  Y_{i,j} | N_i \sim \text{Binomial}(N_i, p).$$

The mean is commonly modeled as a log-linear function of site covariates, as $\text{log}(\lambda_i) = \beta_0 + \beta_1x_i$. Similarly, $p$ is commonly modeled as $\text{logit}(p_{i,j}) = \alpha_0 + \alpha_1x_{i,j}$, a logit-linear function of site-survey covariates.

This estimation approach can be extended to cover $K$ distinct breeding seasons, which correspond with distinct years for wildlife species that breed annually \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. In this case, population closure is assumed across $J$ surveys within year $k$, but is relaxed across years \citep{Kery_Dorazio_Soldaat_Van_Strien_Zuiderwijk_Royle_2009}. A simple specification of a multiple-year model is $N_{i,k} \sim \text{Poisson}(\lambda_{i,k}), \quad Y_{i,j,k} | N_{i,k} \sim \text{Binomial}(N_{i,k}, p_{i,k})$. Like the single-year specification, $\lambda$ is commonly modeled using site and site-year covariates, and $p$ using site-survey-year covariates. There are other variations of N-mixture models that accommodate overdispersed counts through use of a Negative Binomial distribution \citep{Kery_Royle_2010}, a Zero-Inflated Poisson distribution \citep{Wenger_Freeman_2008}, or survey-level random intercepts \citep{Kery_Schaub_2011}. Yet other variations account for non-independent detection probabilities through use of a Beta-Binomial distribution \citep{Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, parse different components of detection through the use of unique covariates \citep{O'Donnell_Thompson_III_Semlitsch_2015}, or relax assumptions of population closure \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}. We do not discuss all of these variations here, but refer interested readers to \cite{Denes_Silveira_Beissinger_2015} for an overview.

The development of metapopulation designs and N-mixture models represents a significant advance in quantitative wildlife ecology. However, there are practical issues that sometimes act as barriers to adoption. Many of the examples of N-mixture models in the wildlife literature have employed Bayesian modeling software such as \proglang{WinBUGS}, \proglang{OpenBUGS}, or \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}. These are extremely powerful and flexible platforms for analyzing hierarchical models, but they come with a few important challenges.  First, many wildlife biologists are not accustomed to coding statistical models using the \proglang{BUGS} modeling syntax. While there are several outstanding resources aimed at teaching this skill \citep{Royle_Dorazio_2008, Kery_2010, Kery_Schaub_2011, Kery_Royle_2015} it is, nonetheless, a considerable commitment. Second, while Markov chain Monte Carlo (MCMC) chains converge quickly for relatively simple N-mixture models, convergence for more complex models can take hours to days, or may not occur at all.

@@

Here we discuss analysis of N-mixture models using the \pkg{R-INLA} package \citep{Martins_Simpson_Lindgren_Rue_2013,Rue_Riebler_Sorbye_Illian_Simpson_Lindgren_2017} for \proglang{R}. The \pkg{R-INLA} package uses integrated nested Laplace approximation (INLA) to derive posterior distributions for a large class of Bayesian statistical models that can be formulated as latent Gaussian models \citep{Rue_Martino_Chopin_2009, Lindgren_Rue_Lindstrom_2011}. INLA was developed to allow estimation of posterior distributions in a fraction of the time taken by MCMC. Like \pkg{unmarked}, the model syntax used in the \pkg{R-INLA} package is a straightforward extension of the modeling syntax commonly used in \proglang{R}. Also, like \pkg{unmarked}, the computational cost of analyzing models with \pkg{R-INLA} is relatively low. The \pkg{R-INLA} approach is different from \pkg{unmarked} in that inference about model parameters falls within a Bayesian framework.

@@

Much of the code used to conduct analyses is shown in the body of this manuscript. However, some repeated code, and code related to generating tables and figures, is not shown for brevity. All code can be accessed via \url{https://github.com/tmeeha/inlaNMix}. See \url{https://r-inla.org} to connect with the community around the development of \pkg{R-INLA} and its application to geostatistics, biostatistics, epidemiology, and econometrics \citep{Lindgren_Rue_2015,Blangiardo_Cameletti_2015}.

@@

As is commonly the case, we assume that the counts are overdispersed due to the effects of unknown variables. Overdispersion was generated and modeled using a Negative Binomial distribution for the count component of the model. The specific model parameters and their simulated values were: p.b0 = 1.0 (logit($p$) intercept), p.b1 = -2.0 (effect of covariate 1 on logit($p$)), p.b3 = 1.0 (effect of covariate 3 on logit($p$)), lam.b0 = 2.0 (log(lambda) intercept), lam.b1 = 2.0 (effect of covariate 1 on log(lambda)), lam.b2 = -3.0 (effect of covariate 2 on log(lambda)), lam.b4 = 1.0 (effect of year on log(lambda)), disp.size = 3.0 (size of the overdispersion parameter). All independent variables in the simulation were centered at zero to alleviate computational difficulties and to make model intercepts more easily interpreted.

@@

\subsection[Analysis with JAGS]{Analysis with JAGS}
We first analyzed the simulated data using \proglang{JAGS}, via the \pkg{runjags} package. In defining the model, we specified a Negative Binomial distribution for the abundance component, and used vague normal priors for the intercepts and the global effects of the covariates of $\lambda$ and $p$.

@@

The first argument in the call to the pcount() function was the model formula, which specified the covariates for detection and then the covariates for abundance. This was followed by an argument identifying the appropriate unmarked frame, and the form of the mixture model, Negative Binomial-Binomial.

@@

The call to the inla() function includes several components. First is the model statement. On the left side of the formula is the counts.and.count.covs object that includes the matrix of counts and the covariates related to $\lambda$.  On the right side of the formula is a 1 to specify a global intercept for $p$ and the two covariates for $p$.  Note that a wide range of random effects for $p$ could be added to the right side of the formula using the f() syntax \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013}. The second argument describes the data, provided here as a list that corresponds with the model formula. Third is the likelihood family, which can take values of "nmix" for a Poisson-Binomial mixture and "nmixnb" for a Negative Binomial-Binomial mixture. The fourth (control.fixed, for detection parameters) and fifth (control.family, for abundance and overdispersion parameters) arguments specify the priors for the two model components. In this case, the priors are specified similarly to those in the \proglang{JAGS} model, but a variety of other options are available. Several other characteristics of the analysis can be modified in the call to inla(). See \cite{Rue_Martino_Lindgren_Simpson_Riebler_2013} for details.

@@

%% discussion
\section[Discussion]{Discussion}
The purpose of this study was to demonstrate the use of the \pkg{R-INLA} package \citep{Rue_Martino_Lindgren_Simpson_Riebler_2013} to analyze N-mixture models and to compare performance of \pkg{R-INLA} to two other common approaches, \proglang{JAGS} \citep{plummer2003jags,Lunn_Jackson_Best_Thomas_Spiegelhalter_2012}, via the \pkg{runjags} package \citep{Denwood_2016}, which employs MCMC methods and allows Bayesian inference, and the \pkg{unmarked} package \citep{Fiske_Chandler_others_2011}, which uses Maximum Likelihood and allows Frequentist inference.

@@

Limitations of \pkg{R-INLA} are mainly related to the more restricted set of N-mixture models that can be specified. Of the approaches explored here, \proglang{BUGS}-based approaches allow users ultimate flexibility in specifying models. For example, with \proglang{JAGS}, site-survey-year covariates for detection are possible, multiple types of mixed distributions are available \citep{Joseph_Elkin_Martin_Possingham_2009,Martin_Royle_Mackenzie_Edwards_Kery_Gardner_2011}, and a variety of random effects can be specified for both $\lambda$ and $p$ \citep{Kery_Schaub_2011}. In comparison, the current version of \pkg{R-INLA} does not handle site-survey covariates, employs only Poisson-Binomial and Negative Binomial-Binomial mixtures, and handles random effects (exchangeable, spatially and temporally structured) for $p$ only. In cases where site-survey covariates are particularly important and not otherwise controlled in the sampling design, \pkg{R-INLA} will not be the appropriate tool (Fig. \ref{fig:fig2}).

When compared to \pkg{unmarked}, the \pkg{R-INLA} approach is similar in regards to model syntax and data format. The approaches are also similar in that both yield results much faster than MCMC. The two approaches differ in that \pkg{R-INLA} is roughly 10 times faster than \pkg{unmarked}, likely due to the different approach used to compute model likelihoods (see Appendix A). They also differ in that \pkg{unmarked} can accommodate site-survey covariates and can analyze dynamic N-mixture models \citep{Chandler_Royle_King_2011, Dail_Madsen_2011}.

@@

%% appendix
\appendix
\section{Recursive computations of the "nmix" likelihood}
The likelihood for the simplest case is
\begin{displaymath}
    \text{Probability}(y) = \sum_{n = y}^{\infty}
    \text{Poisson}(n ; \lambda) \;\times\; \text{Binomial}(y;  n, p)
\end{displaymath}
where $\text{Poisson}(n; \lambda)$ is the density for the Poisson distribution with mean $\lambda$, $\lambda^{n}\exp(-\lambda)/n!$, and $\text{Binomial}(y; n, p)$ is the density for the Binomial distribution with $n$ trials and probability $p$, ${n \choose y} p^{y}(1-p)^{n-p}$. Although the likelhood can be computed directly when replacing the infinite limit with a finite value, we will demonstrate here that we can easily evaluate it using a recursive algorithm that is both faster and more numerical stable. The same idea is also applicable to the Negative Binomial case, and the case where we have replicated observations of the same $n$. We leave it to the reader to derive these straight forward extensions.

The key observation is that both the Poisson and the Binomial distribution can be evaluated recursively in $n$,
\begin{displaymath}
    \text{Poisson}(n; \lambda) = \text{Poisson}(n-1; \lambda) \frac{\lambda}{n}
\end{displaymath}
and
\begin{displaymath}
    \text{Binomial}(y; n, p) = \text{Binomial}(y; n-1, p) \frac{n}{n-y}(1-p),
\end{displaymath}
and then also for the Poisson-Binomial product
\begin{displaymath}
    \text{Poisson}(n ; \lambda) \; \text{Binomial}(y;  n, p)
    =
    \text{Poisson}(n-1; \lambda) \; \text{Binomial}(y; n-1, p)
    \frac{\lambda}{n-y}(1-p).
\end{displaymath}
If we define $f_i = \lambda(1-p)/i$ for $i=1, 2, \ldots$, we can make use of this recursive form to express the likelihood with a finite upper limit as
\begin{eqnarray}
    \text{Prob}(y) &=& \sum_{n = y}^{n_{\text{max}}}
                       \text{Poisson}(n ; \lambda)\;
                      \text{Binomial}(y;  n, p) \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1 + f_1f_2 +
                    \ldots
                      +f_1\cdots f_{n_\text{max}}
                       \Big\} \nonumber\\
                   &=& \text{Poisson}(y; \lambda)\; \text{Binomial}(y; y, p)
                       \Big\{ 1 + f_1(1+f_2(1+f_3(1+ \dots)))\Big\}\nonumber
\end{eqnarray}
The log-liklihood can then be evaluated using the following simple \proglang{R}-code.
\begin{CodeInput}
R> fac <- 1; ff <- lambda * (1-p)
R> for(i in (n.max - y):1){
+    fac <- 1 + fac * ff / i
+  } 
R> log.L <- dpois(y, lambda, log = TRUE) + dbinom(y, y, p, log = TRUE) + 
+           log(fac)
\end{CodeInput}
Since this evaluation is recursive in decreasing $n$, we have to choose the upper limit $n_\text{max}$ up-front, for example as an integer larger than $y$ so that $\frac{\lambda (1-p)}{n_\text{max}-y}$ is small. Note that we are computing \code{fac} starting with the smallest contributions, which are more numerically stable.
